The Power of Unbiased Recursive Partitioning:\\ A Unifying View of CTree, MOB, and GUIDE

### 2020-02-27 pkg/partykit/inst/guideliketest/DAGStat2019/slides.Rnw
added overview slide for GUIDE/CTree/MOB; omit full factorial results
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2990 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: c9a2fc377370b72cb0f452ce43082b6efa818f8b

\title{The Power of Unbiased Recursive Partitioning:\\
A Unifying View of CTree, MOB, and GUIDE}
\author{Lisa Schlosser, Torsten Hothorn, Achim Zeileis}
\setbeamerfont{url}{size*={11.5pt}{13pt},series=\mdseries}
%\renewcommand{\mysize}{\fontsize{7.7pt}{9pt}\selectfont}
\URL{http://www.partykit.org/} 
\headerimage{3}

@@

%% colors
\definecolor{HighlightOrange}{rgb}{0.9490196,0.5725490,0.0000000}
\definecolor{HighlightBlue}{rgb}{0.4784314,0.7490196,0.9803922}

@@

\section{Unbiased Recursive Partitioning}

@@

\begin{frame}%[fragile]
\frametitle{Unbiased Recursive Partitioning}
{\bf Basic tree algorithm:}
\medskip
\begin{enumerate}
\item Fit a model $\mathcal{M}(Y,X;\hat{\beta})$ to the response $Y$ and possible covariates $X$.
\item Assess association of $\mathcal{M}(Y,X;\hat{\beta})$
and each possible split variable~$Z_j$ and select the split~variable~$Z_{j^\ast}$ showing the strongest association.
\item Choose the corresponding split point leading to the highest improvement of model fit and split the data.
\item Repeat steps 1--3 recursively in each of the resulting subgroups until some stopping criterion is met.
\end{enumerate}

@@

\begin{frame}%[fragile]
\frametitle{Split Variable Selection}

@@

\begin{frame}%[fragile]
\frametitle{Discrepancy Measures}

@@

@@

\visible<2>{
\vspace{-0.3cm}
$$\hspace{-0.3cm}{\color{red}\Downarrow} \hspace{3.4cm} {\color{red}\Downarrow} $$
$$
{\color{red} -2 \cdot r(Y, X, \hat{\beta_0},\hat{\beta_1})}  \hspace{0.7cm} {\color{red} -2 \cdot r(Y, X, \hat{\beta_0},\hat{\beta_1}) \cdot X}
$$
}
\end{frame}

\begin{frame}
\frametitle{A Unifying View}

@@

\begin{frame}%[fragile]
\frametitle{General Framework}

@@

\begin{itemize}
\item \emph{Pre-pruning:}
Internal stopping criterion based on Bonferroni-corrected $p$-values of the applied tests. Stop splitting when there is no significant association. 
\item \emph{Post-pruning:}
First grow a very large tree and afterwards prune splits that do not improve the model fit, 
either via cross-validation (e.g.,~cost-complexity pruning as in CART) or 
based on information criteria (e.g.,~AIC or BIC). 
\end{itemize}
\end{frame}

@@

\begin{frame}%[fragile]
\frametitle{Simulation 1: Full factorial analysis of building blocks}

\begin{figure}%[t]
\setkeys{Gin}{width=0.8\linewidth}
<<3way, fig=TRUE, echo=FALSE, height=4, width=8>>=
xyplot(pval_z1 ~ cat | xi, 
       groups = interaction(bin, res_scores), 
       data = sim_3way, 
       ylim = c(0,0.5),
       type = c("a"), 
       col = c(pal[1], pal[3], pal[1], pal[3]),
       lty = c(1,1,2,2),
       lwd = 2,
       layout = c(2,1),
       scales = list(alternating=FALSE),
       xlab = "Categorization",
       ylab = expression("p-value of"~Z[1]),
       strip = strip.custom(factor.levels = c(expression(xi==0~"(50%)"~" and"~delta==0.3), 
                                              expression(xi==0.8~"(90%)"~" and"~delta==1)),
                            bg = "gray90"),
       key = list(text = list(c("Goodness of fit:", "  residuals", "  scores", "Dichotomization:", "  yes", "  no")),
                  columns = 2,
                  lines = list(lty = c(0,1,2,0,1,1), 
                               type = "l", lwd = 2,
                               col = c("black","black","black","black", pal[1], pal[3]))))

@@

@@

\begin{itemize}
\setlength\itemsep{0.2cm}
\item Full model scores better than residuals only.
\item Original values of scores/residuals better than binarized values.
\item Categorization is simpler, but less powerful in margins.
\item If the significance test performs well pre-pruning works well,\\
otherwise post-pruning might be needed.
%\item Including full model scores is advantageous over using only residuals.
%\item Binarizing scores/residuals does not improve the performance.
%\item Categorizing split variables can simplify calculations, however,
%effects might be missed unless a binning point is close to the true split point.
%\item If the applied test performs well as significance test pre-pruning
%is preferable, otherwise post-pruning.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Software}

\textbf{Package:} Available on CRAN at

\url{https://CRAN.R-project.org/package=partykit}\\

\textbf{Replication code:} Available on R-Forge at

\url{https://R-Forge.R-project.org/projects/partykit/}\\

%\bigskip

@@

%\begin{itemize}
%\item \texttt{ctree}
%\item \texttt{mob}
%\item Adaptions of both of them
%\item Reimplementation of GUIDE
%\end{itemize}
\end{frame}

@@

\footnotesize
Schlosser L, Hothorn T, Zeileis A (2019).
 \dquote{The Power of Unbiased Recursive Partitioning: A Unifying View of CTree, MOB, and GUIDE.}
 \emph{Working paper in preparation}.

Breiman L, Friedman J H, Olshen R A, Stone C J (1984).
 \textit{Classification and Regression Trees.}
 Wadsworth,
 California.

@@

Hothorn T, Zeileis A (2015).
 \dquote{\textbf{partykit}: A Modular Toolkit for Recursive Partytioning in \textsf{R}.}
 \textit{Journal of Machine Learning Research},
 \textbf{16}, 3905--3909.
 \url{http://www.jmlr.org/papers/v16/hothorn15a.html}

>>>>

\title{The Power of Unbiased Recursive Partitioning:\\
A Unifying View of CTree, MOB, and GUIDE}
\author{Lisa Schlosser, Torsten Hothorn, Achim Zeileis}
\setbeamerfont{url}{size*={11.5pt}{13pt},series=\mdseries}
%\renewcommand{\mysize}{\fontsize{7.7pt}{9pt}\selectfont}
\URL{http://www.partykit.org/partykit} 
\headerimage{4}

@@

%% colors
\definecolor{HighlightOrange}{rgb}{0.9490196,0.5725490,0.0000000}
\definecolor{HighlightBlue}{rgb}{0.4784314,0.7490196,0.9803922}
\definecolor{dred}{RGB}{119,16,26}

@@

\section{The Power of Unbiased Recursive Partitioning}

\subsection{Motivation}

@@

\subsection{Unbiased recursive partitioning}

\begin{frame}
\frametitle{Unbiased recursive partitioning}

\textbf{GUIDE:} Loh (2002, \emph{Statistica Sinica}).
\begin{itemize}
  \item First \emph{unbiased} algorithm for recursive partitioning of linear models.
  \item Separation of split variable and split point selection.
  \item Based on $\chi^2$ tests.
\end{itemize}

\textbf{CTree:} Hothorn, Hornik, Zeileis (2006, \emph{JCGS}).
\begin{itemize}
  \item Proposed as unbiased recursive partitioning for nonparametric modeling.
  \item Based on conditional inference (or permutation tests).
  \item Can be model-based via model scores as the response transformation.
\end{itemize}

\textbf{MOB:} Zeileis, Hothorn, Hornik (2008, \emph{JCGS}).
\begin{itemize}
  \item Model-based recursive partitioning using M-estimation (ML, OLS, CRPS, \dots).
  \item Based on parameter instability tests.
  \item Adapted to various psychometric models: Rasch, PCM, Bradley-Terry, MPT, SEM, networks, \dots.
\end{itemize}

\begin{frame}
\frametitle{Unbiased recursive partitioning}
{\bf Basic tree algorithm:}
\medskip
\begin{enumerate}
\item Fit a model $\mathcal{M}(Y,X;\hat{\beta})$ to the response $Y$ and possible covariates $X$.
\item Assess association of $\mathcal{M}(Y,X;\hat{\beta})$
and each possible split variable~$Z_j$ and select the split~variable~$Z_{j^\ast}$ showing the strongest association.
\item Choose the corresponding split point leading to the highest improvement of model fit and split the data.
\item Repeat steps 1--3 recursively in each of the resulting subgroups until some stopping criterion is met.
\end{enumerate}

@@

\begin{frame}%[fragile]
\frametitle{Split variable selection}

@@

\begin{frame}%[fragile]
\frametitle{Discrepancy measures}

@@

\bigskip
\pause

@@

\pause

\vspace{-0.3cm}
$$\hspace{-0.3cm}{\color{dred}\Downarrow} \hspace{3.4cm} {\color{dred}\Downarrow} $$
$$
{\color{dred} -2 \cdot r(Y, X, \hat{\beta_0},\hat{\beta_1})}  \hspace{0.7cm} {\color{dred} -2 \cdot r(Y, X, \hat{\beta_0},\hat{\beta_1}) \cdot X}
$$

\subsection{A unifying view}

\begin{frame}
\frametitle{A unifying view}

@@

\begin{frame}%[fragile]
\frametitle{General framework}

@@

\begin{itemize}
\item \emph{Pre-pruning:}
Internal stopping criterion based on Bonferroni-corrected $p$-values of the underlying tests. Stop splitting when there is no significant association. 
\item \emph{Post-pruning:}
First grow a very large tree and afterwards prune splits that do not improve the model fit, 
either via cross-validation (e.g.,~cost-complexity pruning as in CART) or 
based on information criteria (e.g.,~AIC or BIC). 
\end{itemize}
\end{frame}

\subsection{Simulation}

@@

%% \begin{frame}%[fragile]
%% \frametitle{Simulation 1: Full factorial analysis of building blocks}
%% 
%% \begin{figure}%[t]
%% \setkeys{Gin}{width=0.8\linewidth}
%% <<3way, fig=TRUE, echo=FALSE, height=4, width=8>>=
%% xyplot(pval_z1 ~ cat | xi, 
%%        groups = interaction(bin, res_scores), 
%%        data = sim_3way, 
%%        ylim = c(0,0.5),
%%        type = c("a"), 
%%        col = c(pal[1], pal[3], pal[1], pal[3]),
%%        lty = c(1,1,2,2),
%%        lwd = 2,
%%        layout = c(2,1),
%%        scales = list(alternating=FALSE),
%%        xlab = "Categorization",
%%        ylab = expression("p-value of"~Z[1]),
%%        strip = strip.custom(factor.levels = c(expression(xi==0~"(50%)"~" and"~delta==0.3), 
%%                                               expression(xi==0.8~"(90%)"~" and"~delta==1)),
%%                             bg = "gray90"),
%%        key = list(text = list(c("Goodness of fit:", "  residuals", "  scores", "Dichotomization:", "  yes", "  no")),
%%                   columns = 2,
%%                   lines = list(lty = c(0,1,2,0,1,1), 
%%                                type = "l", lwd = 2,
%%                                col = c("black","black","black","black", pal[1], pal[3]))))
%% 
%% @
%% \end{figure}
%% \end{frame}

@@

\subsection{Recommendations}

@@

\begin{itemize}
\setlength\itemsep{0.2cm}
\item Full model scores better than residuals only.
\item Original values of scores/residuals better than binarized values.
\item Categorization is simpler, but less powerful in margins.
\item Maximally-selected statistics (as in MOB) more powerful for abrupt shifts.
\item Linear statistics (default in CTree) more powerful for linear changes.
\item If the significance tests perform well pre-pruning works well,\\
      otherwise post-pruning might be needed.
\end{itemize}
\end{frame}

%% \begin{frame}
%% \frametitle{Software}
%% 
%% \textbf{Package:} Available on CRAN at
%% 
%% \url{https://CRAN.R-project.org/package=partykit}\\
%% 
%% \bigskip
%% 
%% \textbf{Replication code:} Available on R-Forge at
%% 
%% \url{https://R-Forge.R-project.org/projects/partykit/}\\
%% 
%\bigskip

@@

%\begin{itemize}
%\item \texttt{ctree}
%\item \texttt{mob}
%\item Adaptions of both of them
%\item Reimplementation of GUIDE
%\end{itemize}
%% \end{frame}

@@

\footnotesize
Schlosser L, Hothorn T, Zeileis A (2019).
  \dquote{The Power of Unbiased Recursive Partitioning: A Unifying View of CTree, MOB, and GUIDE.}
  arXiv:1906.10179, \emph{arXiv.org E-Print Archive}.
  \url{https://arxiv.org/abs/1906.10179}.

\bigskip
\bigskip

@@
